---
title: "ECL-298"
author: "Yuri Victor Calvo"
date: "`r Sys.Date()`"
output: html_document
---

# Packages

```{r include=F}
library(tidyverse)
library(janitor)
library(maps)
library(randomForest)
library(pROC)
```

# Df Import

The `farms` Dataset was provided by the California Department of Fishery and Agriculture, it contains information from all poultry farms tested for HPAI from 2022 until February of 2025.

Relevant variables for the project:

-   ID

-   Diagnosis: positive or negative for HPAI

-   start_date: Date in which the farm was tested

-   species_gp: What animal species was tested.

The `df` Dataset, was obtained with Google Earth Engine (GEE), from the GridMET projections. It contains Weather information of all poultry farms from the farms dataset, 14 prior and during the day they were tested. The information refers to a 2km buffer along the farm.

Relevant variables for the project:

-   temp_mean: Mean temperature for the day

-   rh_mean: mean Relative Humidity for the day

-   cpr: accumulated preciptation

-   ref: day reference; 0=test date

-   wave: outbreak wave

-   county: county where the farm is located

```{r include=F}
pathF<- "C:/Users/yuriv/Box/MPVM (Yuri Calvo)/Spring 2025/MPM 204/Farms_County.csv"
farms <- read_csv(pathF)
farms <- clean_names(farms)
farms <- farms %>% 
  filter(species_gr %in% c("Chicken","A - Select Species Group","Poultry"))
farms$month <- ymd(paste(year(farms$date),month(farms$date),'01',sep="-"))
path <- "C:/Users/yuriv/Box/MPVM (Yuri Calvo)/Final Project/Data/DBs/treated/GRIDMET/2-2_weeks_col_treat/GRIDMET_10km_treat.csv"

tests_dates <- farms %>% dplyr::select(project_id,start_date)

df <- read_csv(path)
df <- clean_names(df)
df <- df %>% 
  filter(ref<=0 & id %in% farms$project_id) %>% 
  mutate(ref=ref*(-1)) 


farms$id <- farms$project_id

df <- left_join(df,farms,by="id")

df$diag <- ifelse(df$diagnosis=="Positive",1,0)
df$waves <- as.factor(df$waves)
df <- df %>% 
  dplyr::select(id,date.x,ref,cpr,rh_mean,temp_mean,diag,name,waves,latitude,longitude) %>% 
  filter(ref>=3)

colnames(df) <- c("id",'date','ref','cpr','rh_mean','temp_mean','diag','county','waves','lat','long')
#######

rm(path,pathF)


```

## Spatio-Temporal Distribution

The plot below, indicates a clear seasonality in the cases of HPAi in California, with outbreaks ceasing during summer. The seasonality of HPAI is widely studied, and attributed to birds migration routes, as well and weather conditions that affect where they land.

```{r}
posi <- farms %>% 
  filter(diagnosis=="Positive") %>% 
  group_by(month) %>% 
  summarise(count=n(),waves=unique(waves))
  
farms %>% 
    group_by(waves,diagnosis) %>% 
    summarise(n=n())
  

ggplot(posi,aes(x=month,y=count))+
  geom_bar(mapping=aes(fill=as.factor(waves)),stat="identity")+
  scale_x_date(date_breaks="3 months",date_labels = "%m/%Y")+
  labs(title = "Farms positive for HPAI in California from 2022-2025",
       x="Date",
       y="Count",
       fill="Outbreak Waves")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))
  
```

Similarly to the wave distribution, there is also a spatial effect present in the cases as well. By observing the map below we can see that the distribution of cases across the state is very uneven between counties.

```{r}
CA <- map_data('county','california')

ggplot()+
  geom_map(CA,map = CA,mapping=aes(,map_id=region),colour="black",fill="white")+
  geom_point(subset(df,diag==1 & ref==3),mapping=aes(x=long,y=lat,colour=waves),size=1,position =position_jitter(0.05,0.1))+
  expand_limits(x=CA$long,y=CA$lat)+
  labs(title="HPAI cases in California")
```

# Data prep

```{r}
df_3d <- df %>% 
  mutate(dg=case_when(
      ref <= 5  ~ "5_7",
      ref <= 10  ~ "8_10",
      ref <= 12 ~ "10_12",
      ref <= 15 ~ "13_14")) %>% 
  group_by(id,dg,diag,waves) %>%
  summarise(temp=mean(temp_mean),rh=mean(rh_mean),cpr=mean(cpr),.groups = "drop") %>% 
  pivot_wider(id_cols = c(id,diag,waves),names_from = dg,values_from = c(temp,rh,cpr),names_glue ="{.value}_{dg}")




```

# Cross Validation

```{r}
n=floor(NROW(df_3d)/3)
i=sample(NROW(df_3d),n)
train <- df_3d[i,]
test <- df_3d[-i,]
```

# Random Forests

The random forest indicates that out of preciptation, relative humidity and temperature, temperature is the most important variable, followed by rh and cpr.

Is worth noting that this is an analysis made from the mean values from from the period of time to avoid inflated values.

# 3 days ranges

```{r}
m <- randomForest(diag~.-waves-id,data=train,ntree=5000)
varImpPlot(m,main="Importance Plot")
```

## Prediction

This is the result of the random forest analysis when compared to the actual data.

Quite impressive results, Likely influenced by the imbalance between negative and positive cases.

The ROC curve also indicates some potential issue with the data.

Another potential source, that needs to be checked is the cpr variable. Its right skewed and extremely zero inflated, so it can negatively affect the observed values.

```{r}
pred_train <- as.numeric(predict(m,train))-1
table(pred_train,train$diag)

r <- roc(train$diag,pred_train, levels = c("0", "1"))

plot.roc(r,legacy.axes = T,print.auc = T,main="Train Data")
```

```{r}
pred_test<- as.numeric(predict(m,test))-1
table(pred_test,test$diag)

r <- roc(test$diag,pred_test)

plot.roc(r,legacy.axes = T,print.auc = T,main="Test Data")

```

# GLM

```{r}
mod_rf <- glm(diag~temp_10_12+rh_13_14+cpr_10_12,data=train,family=binomial)
summary(mod_rf)

```

```{r}
pred_train_rf <-  predict(mod_rf,train,type = "response")

rrf <- roc(train$diag,pred_train_rf)

plot.roc(rrf,legacy.axes = T,print.auc = T,main="Train Data Most Important")

```

```{r}
pred_test_rf <-  predict(mod_rf,test,type = "response")
rrf <- roc(test$diag,pred_test_rf)

plot.roc(rrf,legacy.axes = T,print.auc = T,main="Test Data Most Important")




```
